%% ============================================================
%%  Deep Learning Course — Session 01
%%  Topic : Deep Learning Introduction & Course Plan
%%  Deck  : 01DLPlan (21 slides, November 17 2025)
%%  Instructors: Prof. S R M Prasanna & Dr. Sunil Saumya
%%  Institution: IIIT Dharwad
%% ============================================================
\documentclass[12pt,a4paper]{article}

%% ---------- mathematics ----------
\usepackage{amsmath,amsthm,amssymb,mathtools}

%% ---------- graphics & diagrams ----------
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,calc,fit,backgrounds}
\usepackage{graphicx}

%% ---------- tables ----------
\usepackage{booktabs,tabularx,multirow}

%% ---------- algorithms ----------
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}

%% ---------- coloured boxes ----------
\usepackage[most]{tcolorbox}

%% ---------- hyperlinks & cross-refs ----------
\usepackage{hyperref}
\usepackage[capitalize,noabbrev]{cleveref}

%% ---------- page layout ----------
\usepackage[margin=1in]{geometry}
\usepackage{fancyhdr}
\usepackage{enumitem}
\usepackage{xcolor}

%% ============================================================
%%  Custom tcolorbox environments
%% ============================================================
\newtcolorbox{keyinsight}[1][]{%
  colback=blue!5!white,colframe=blue!75!black,
  fonttitle=\bfseries,title=Key Insight,#1}

\newtcolorbox{example}[1][]{%
  colback=green!5!white,colframe=green!75!black,
  fonttitle=\bfseries,title=Example,#1}

\newtcolorbox{warningbox}[1][]{%
  colback=red!5!white,colframe=red!75!black,
  fonttitle=\bfseries,title=Warning,#1}

%% ============================================================
%%  amsthm environments
%% ============================================================
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}{Definition}[section]
\newtheorem{remark}{Remark}[section]

%% ============================================================
%%  Page style
%% ============================================================
\pagestyle{fancy}
\fancyhf{}
\rhead{Deep Learning — Session 01}
\lhead{IIIT Dharwad}
\rfoot{Page \thepage}
\lfoot{Prof.\ S R M Prasanna \& Dr.\ Sunil Saumya}

%% ============================================================
%%  Metadata
%% ============================================================
\title{%
  \vspace{-1cm}
  {\Large\bfseries Deep Learning Course}\\[4pt]
  {\large Session 01: Introduction, Course Plan, and\\
   Historical Foundations of Deep Learning}\\[6pt]
  {\normalsize IIIT Dharwad --- M.Tech Programme}
}
\author{%
  Prof.\ S R M Prasanna \quad \& \quad Dr.\ Sunil Saumya\\[4pt]
  \small Lecture Notes compiled from Video (Vimeo) and Slide Deck \texttt{01DLPlan}
}
\date{November 17, 2025}

%% ============================================================
\begin{document}
%% ============================================================

\maketitle
\thispagestyle{fancy}

\begin{abstract}
These notes document Session~01 of the Deep Learning course offered at
IIIT Dharwad to M.Tech students across super-specialisations.  The session
serves as a course orientation and motivation lecture: it presents the
8-week course plan with PyTorch hands-on components, traces the history of
deep learning from the 2006 Hinton \emph{et al.}\ breakthrough, explains the
conceptual distinction between classical Machine Learning and Deep Learning,
and situates Deep Learning within the broader AI taxonomy
(AI~$\supset$~ML~$\supset$~NN~$\supset$~DL~$\supset$~GAI).  Mathematical
notation, TikZ diagrams, and structured summaries are included to aid
self-study.
\end{abstract}

\tableofcontents
\newpage

%% ============================================================
\section{Introduction}
\label{sec:intro}
%% ============================================================

Deep Learning is described by the instructors as a \emph{cornerstone course}
for Data Science and Artificial Intelligence.  The course is mandatory for all
M.Tech super-specialisations because modern intelligent systems — from
automatic speech recognition to large language models — rely on deep neural
architectures.

The primary pedagogical shift in this offering, compared to earlier editions of
the course, is the integration of \emph{hands-on PyTorch programming} alongside
each theory module.  As Prof.\ Prasanna explains in the lecture:

\begin{quote}
``Since you are all working professionals, we thought: how do we give
maximum benefit of this course?  Dr.\ Sunil suggested we make this course
more hands-on, such that once you complete it you also have exposure to open
source toolkits for training and testing deep learning models, which will be
useful from a career point of view.''
\end{quote}

\subsection{Session Objectives}
\label{subsec:objectives}

By the end of Session~01 the student should be able to:
\begin{enumerate}[label=\textbf{O\arabic*.}]
  \item Identify the eight-week syllabus structure and the corresponding
        hands-on PyTorch tasks.
  \item Describe the 2006 Hinton \emph{et al.}\ breakthrough and explain why
        it revived the deep learning field.
  \item Distinguish Classical Machine Learning from Deep Learning in terms
        of feature engineering.
  \item Situate Deep Learning within the AI~$\supset$~ML~$\supset$~NN~$\supset$~DL~$\supset$~GAI hierarchy.
  \item Appreciate the three enabling factors that made deep learning
        practical: (i) algorithmic advances, (ii) big data availability, and
        (iii) GPU computing power.
\end{enumerate}

\subsection{Course Metadata}
\label{subsec:metadata}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Item} & \textbf{Detail} \\
\midrule
Course name   & Deep Learning (3 Credits) \\
Semester      & Second Semester, M.Tech \\
Instructors   & Prof.\ S R M Prasanna; Dr.\ Sunil Saumya \\
Slide deck    & \texttt{01DLPlan} (21 slides) \\
Video         & \url{https://vimeo.com/1137772716} \\
Platform      & PyTorch (open-source) \\
Assessment    & Class notes 15\%; X-factor 10\%; Assignments 20\%;\\
              & Quiz 10\%; Mid-term; End-term (65\% written total) \\
\bottomrule
\end{tabular}
\end{center}

%% ============================================================
\section{Background and Prerequisites}
\label{sec:background}
%% ============================================================

\subsection{Recommended Reading}
\label{subsec:books}

The instructors provided the following reference list for students who wish to
go beyond the lecture slides.  These texts are \emph{supplementary}; the
slide decks and hands-on notebooks are self-sufficient for the course.

\begin{enumerate}[label=\arabic*.]
  \item R.\ O.\ Duda and P.\ E.\ Hart, \emph{Pattern Classification},
        Wiley, 2001.
  \item C.\ M.\ Bishop, \emph{Neural Networks for Pattern Recognition},
        Oxford University Press, 1995.
  \item B.\ Yegnanarayana, \emph{Artificial Neural Networks},
        PHI Learning, 1999.
  \item C.\ M.\ Bishop, \emph{Pattern Recognition and Machine Learning (PRML)},
        Springer, 2006.
  \item M.\ Nielsen, \emph{Neural Networks and Deep Learning} (open book,
        freely available online).
  \item I.\ Goodfellow \emph{et al.}, \emph{Deep Learning},
        MIT Press, 2016.
  \item C.\ C.\ Aggarwal, \emph{Neural Networks and Deep Learning},
        Springer, 2018.
  \item M.\ M.\ Khapra, \emph{Deep Learning Parts I \& II},
        NPTEL Course (IIT Madras).
  \item A.\ Ghodsi, \emph{Deep Learning},
        University of Waterloo (online lectures).
  \item C.\ M.\ Bishop and H.\ Bishop, \emph{Deep Learning},
        Springer, 2024.
\end{enumerate}

\begin{keyinsight}
Prof.\ Prasanna personally recommends the NPTEL course by Mithesh Khapra
(IIT Madras) and Ali Ghodsi's University of Waterloo course as complementary
online resources.  The 2024 Bishop \& Bishop text is also highlighted as the
most up-to-date reference on the subject.
\end{keyinsight}

\subsection{Mathematical Prerequisites}
\label{subsec:mathprereq}

Students are expected to be comfortable with the following:

\begin{itemize}
  \item \textbf{Linear algebra}: matrix multiplication, eigenvalues, norms.
  \item \textbf{Calculus}: partial differentiation, chain rule — essential for
        back-propagation.
  \item \textbf{Probability and statistics}: Bayes rule, Gaussian distributions,
        maximum-likelihood estimation.
  \item \textbf{Optimisation}: gradient descent, convexity, saddle points.
  \item \textbf{Programming}: Python fluency; basic familiarity with NumPy and
        Matplotlib.
\end{itemize}

The key mathematical relation underlying supervised learning is a functional
mapping:
\begin{equation}
  \label{eq:functional_mapping}
  \hat{y} = f(\mathbf{x};\,\boldsymbol{\theta}),
\end{equation}
where $\mathbf{x} \in \mathbb{R}^d$ is the input feature vector, $\hat{y}$ is
the predicted output, and $\boldsymbol{\theta}$ denotes all learnable
parameters.  The learning procedure minimises a loss:
\begin{equation}
  \label{eq:loss}
  \mathcal{L}(\boldsymbol{\theta})
  = \frac{1}{N}\sum_{i=1}^{N} \ell\!\left(y_i,\, f(\mathbf{x}_i;\boldsymbol{\theta})\right),
\end{equation}
where $\{(\mathbf{x}_i, y_i)\}_{i=1}^{N}$ is the training set and $\ell$ is a
per-sample loss function (e.g., cross-entropy for classification, mean-squared
error for regression).

%% ============================================================
\section{Course Structure and Syllabus}
\label{sec:courseplan}
%% ============================================================

The 8-week core curriculum is organised so that each session has a \emph{theory
component} (aligned with the official syllabus) and a \emph{hands-on component}
implemented in PyTorch.  \cref{tab:weeks12,tab:weeks34,tab:weeks58} reproduce
the three course-plan slides presented in the lecture.

\subsection{Weeks 1--2: Neural Networks Foundations}
\label{subsec:weeks12}

\begin{table}[ht]
\centering
\caption{Deep Learning Course Plan — Weeks 1--2 (Slide V03).}
\label{tab:weeks12}
\begin{tabularx}{\textwidth}{cclX}
\toprule
\textbf{Week} & \textbf{Session} & \textbf{Theory Topics} & \textbf{Hands-on (PyTorch)} \\
\midrule
1 & 1 & Biological vs.\ Artificial Neuron; Neural Networks; Learning Laws
      & Implement perceptron (logical gates) in PyTorch \\
\addlinespace
1 & 2 & Feedforward Networks; Backpropagation
      & Manual backprop; Build 2-layer NN in PyTorch \\
\addlinespace
2 & 3 & Deep Feedforward NNs; Training Deep Networks; Vanishing Gradients
      & Train deep MLP on MNIST \\
\addlinespace
2 & 4 & Greedy Layer-wise Training; Training Belief Networks; Practical Tricks
      & Layer-wise pre-training demo; effect of depth \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Weeks 3--4: Convolutional Neural Networks}
\label{subsec:weeks34}

\begin{table}[ht]
\centering
\caption{Deep Learning Course Plan — Weeks 3--4 (Slide V04).}
\label{tab:weeks34}
\begin{tabularx}{\textwidth}{cclX}
\toprule
\textbf{Week} & \textbf{Session} & \textbf{Theory Topics} & \textbf{Hands-on (PyTorch)} \\
\midrule
3 & 5 & CNN vs.\ Other NNs; Convolution Basics; Filters; Feature Maps
      & Manual convolution; visualising filters \\
\addlinespace
3 & 6 & CNN Modules: Pooling, Padding, Stride; CNN Training Pipeline
      & Train simple CNN \\
\addlinespace
4 & 7 & CNN as Feature Extractor; CNN as Classifier
      & Use pre-trained CNN for feature extraction \\
\addlinespace
4 & 8 & Significance of CNNs in Object Classification Tasks
      & Error analysis, confusion matrix, Grad-CAM \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Weeks 5--8: Sequence Models and Transformers}
\label{subsec:weeks58}

\begin{table}[ht]
\centering
\caption{Deep Learning Course Plan — Weeks 5--8 (Slide V05).}
\label{tab:weeks58}
\begin{tabularx}{\textwidth}{cclX}
\toprule
\textbf{Week} & \textbf{Session} & \textbf{Theory Topics} & \textbf{Hands-on (PyTorch)} \\
\midrule
5  &  9 & Sequence Models
       & Build simple RNN for sequence prediction \\
\addlinespace
5  & 10 & Recurrent Neural Networks (RNN): Architecture, BPTT
       & RNN-based character/word-level model \\
\addlinespace
6  & 11 & Long Short-Term Memory (LSTM) Networks
       & LSTM for sentiment analysis \\
\addlinespace
6  & 12 & Gated Recurrent Units (GRU)
       & Train GRU and compare with LSTM \\
\addlinespace
7  & 13 & Seq2Seq Models: Encoder--Decoder Architecture
       & Seq2Seq for toy translation / copying task \\
\addlinespace
7  & 14 & Attention Modelling: Bahdanau, Luong
       & Add attention to Seq2Seq model \\
\addlinespace
8  & 15 & Attention: Scaled Dot-Product, Multi-Head
       & Implement scaled dot-product attention \\
\addlinespace
8  & 16 & Transformers: Encoder/Decoder Architecture
       & Mini-transformer block implementation \\
\bottomrule
\end{tabularx}
\end{table}

\begin{keyinsight}
The progression follows a deliberate pedagogical arc:
\begin{center}
Perceptron $\;\longrightarrow\;$ Deep MLP $\;\longrightarrow\;$ CNN
$\;\longrightarrow\;$ RNN/LSTM $\;\longrightarrow\;$ Transformer
$\;\longrightarrow\;$ Generative Models.
\end{center}
Each transition introduces a new inductive bias suited to a different
problem structure (static feature vectors, spatial grids, temporal sequences,
long-range dependencies).
\end{keyinsight}

\subsection{Sequence-to-Sequence and Generative Models (Overview)}
\label{subsec:seq2gen}

After Week~8, the course extends to:
\begin{itemize}
  \item \textbf{Weeks 9--10}: Transformers in detail — architecture, training,
        applications (machine translation, question answering, named entity
        recognition).
  \item \textbf{Weeks 11--13}: Generative models — Variational Autoencoders
        (VAEs), Generative Adversarial Networks (GANs), Generative Pre-trained
        Transformer (GPT).
  \item \textbf{Week 14}: Advanced topics and extended code demonstrations
        (time-permitting).
\end{itemize}

The key concept underpinning sequence-to-sequence learning is the
\emph{encoder--decoder} formulation.  Given an input sequence
$\mathbf{x} = (x_1, x_2, \ldots, x_T)$, the encoder maps it to a fixed-length
context vector $\mathbf{c}$:
\begin{equation}
  \label{eq:encoder}
  \mathbf{c} = \mathrm{Enc}(x_1, \ldots, x_T;\,\boldsymbol{\theta}_\mathrm{enc}),
\end{equation}
and the decoder generates the output sequence
$\mathbf{y} = (y_1, y_2, \ldots, y_{T'})$ one token at a time:
\begin{equation}
  \label{eq:decoder}
  P(y_t \mid y_1,\ldots,y_{t-1},\,\mathbf{c};\,\boldsymbol{\theta}_\mathrm{dec}).
\end{equation}

%% ============================================================
\section{History of Deep Learning}
\label{sec:history}
%% ============================================================

\subsection{The Neural Network Era (Pre-2006)}
\label{subsec:pre2006}

The intellectual lineage of deep learning traces back to artificial neural
networks inspired by the structure and functioning of the human brain.  The
brain achieves its remarkable cognitive power not through a single neuron, but
through \emph{billions of neurons} working in a loosely connected,
parallel, and distributed fashion.  Early researchers sought to mimic this
\emph{parallel and distributed processing} on digital computers.

\begin{definition}[Artificial Neural Network]
\label{def:ann}
An \emph{Artificial Neural Network} (ANN) is a parameterised, layered
computational graph of artificial neurons, each computing a non-linear
function of a weighted sum of its inputs.  Formally, the output of neuron $j$
in layer $l$ is:
\begin{equation}
  \label{eq:neuron}
  a_j^{(l)} = \phi\!\left(\sum_{k} w_{jk}^{(l)}\, a_k^{(l-1)} + b_j^{(l)}\right),
\end{equation}
where $w_{jk}^{(l)}$ are synaptic weights, $b_j^{(l)}$ is a bias,
and $\phi(\cdot)$ is an activation function.
\end{definition}

Shallow neural networks (1--2 hidden layers) emerged as an alternative
computing model for classical machine learning tasks.  They were trained using
the \emph{Back-Propagation} (BP) algorithm, which applies the chain rule to
compute gradients of the loss with respect to all parameters:
\begin{align}
  \label{eq:backprop}
  \frac{\partial \mathcal{L}}{\partial w_{jk}^{(l)}}
  &= \frac{\partial \mathcal{L}}{\partial a_j^{(l)}}
     \cdot \phi'\!\left(z_j^{(l)}\right)
     \cdot a_k^{(l-1)},
\end{align}
where $z_j^{(l)} = \sum_k w_{jk}^{(l)} a_k^{(l-1)} + b_j^{(l)}$ is the
pre-activation.

Despite theoretical appeal, shallow networks could not consistently outperform
classical machine learning methods (SVMs, GMMs, decision trees) on many
benchmarks.  This led researchers to hypothesise that \emph{deeper}
architectures — with more layers and more parameters — would capture richer
representations.

\subsection{The Dark Age: Vanishing Gradients and Insufficient Data}
\label{subsec:darkage}

Attempts to scale up neural networks to many layers ran into two fundamental
obstacles:

\begin{warningbox}
\textbf{Obstacle 1 — Vanishing Gradient Problem.}
When back-propagating through many layers, the gradient signal is multiplied
by the weight matrix and the derivative of the activation function at each
step.  For activations such as the sigmoid $\sigma(z) = 1/(1+e^{-z})$, the
derivative $\sigma'(z) = \sigma(z)(1-\sigma(z)) \leq 0.25$ at all points.
Through $L$ layers, the gradient can shrink exponentially:
\begin{equation}
  \label{eq:vanishing}
  \left\|\frac{\partial \mathcal{L}}{\partial \mathbf{W}^{(1)}}\right\|
  \;\propto\; \prod_{l=2}^{L} \left\|\mathbf{W}^{(l)}\right\|
              \prod_{l=2}^{L} \left|\sigma'\!\left(z^{(l)}\right)\right|
  \;\longrightarrow\; 0 \quad \text{as } L \to \infty.
\end{equation}
Parameters in early layers receive almost no gradient signal and fail to learn.

\textbf{Obstacle 2 — Insufficient Training Data.}
In the 1990s and early 2000s, labelled datasets were small by modern
standards.  A large deep network trained on insufficient data
\emph{over-fits}: it memorises the training examples and fails to generalise
to new inputs.
\end{warningbox}

As a result, most of the research community abandoned deep architectures in
the late 1990s.

\subsection{The 2006 Breakthrough: Hinton \emph{et al.}}
\label{subsec:hinton2006}

A small group of researchers — led by Geoffrey Hinton at the University of
Toronto, who later received both the Turing Award and the Nobel Prize in
Physics for this work — continued to pursue deep architectures through the
early 2000s.  In 2006 they published a landmark paper:

\begin{quote}
Geoffrey E.\ Hinton, Simon Osindero, and Yee-Whye Teh,
``A Fast Learning Algorithm for Deep Belief Nets,''
\emph{Neural Computation}, vol.\ 18, pp.\ 1527--1554, 2006.
\end{quote}

\begin{keyinsight}[title=The 2006 Breakthrough]
Hinton \emph{et al.}\ introduced \emph{greedy layer-wise pre-training}:
each layer of a Deep Belief Network (DBN) is first trained unsupervised as a
Restricted Boltzmann Machine (RBM), using efficient contrastive divergence.
The pre-trained weights then serve as a good initialisation for fine-tuning
the entire network with back-propagation.  This strategy:
\begin{enumerate}
  \item Circumvented the vanishing gradient problem by initialising weights
        near a good solution.
  \item Achieved state-of-the-art accuracy ($>98\%$) on the MNIST handwritten
        digit recognition benchmark.
  \item Demonstrated that deep networks are trainable and outperform
        classical ML on real tasks.
\end{enumerate}
This paper gave deep learning its modern name and triggered a ``snowball
effect'' that, within a decade (2006--2016), led to deep learning conquering
industry benchmarks.
\end{keyinsight}

\subsection{The Three Enabling Factors (Post-2006)}
\label{subsec:enablers}

The instructor identifies three converging advances that made modern deep
learning possible:

\begin{enumerate}[label=\textbf{E\arabic*.}]
  \item \textbf{Algorithmic innovations}: greedy pre-training, ReLU
        activations, Dropout, Batch Normalisation, residual connections.
  \item \textbf{Big data}: the connected world generated vast labelled
        datasets (ImageNet, Common Crawl, LibriSpeech, etc.).
  \item \textbf{GPU computing}: graphics processors enabled massively parallel
        matrix multiplications, reducing training time from weeks to hours.
\end{enumerate}

\begin{remark}
\label{rem:timeline}
The snowball accelerated rapidly.  AlexNet (2012) won ImageNet with a
top-5 error of $15.3\%$, crushing the classical-vision runner-up at $26.2\%$.
By 2016 deep learning held state-of-the-art on image classification, speech
recognition, machine translation, and game-playing.  The introduction of the
Transformer architecture (Vaswani \emph{et al.}, 2017) and subsequent
large language models further accelerated the field into the Generative AI era.
\end{remark}

%% ============================================================
\section{Machine Learning vs.\ Deep Learning}
\label{sec:mldl}
%% ============================================================

\subsection{Classical Machine Learning Pipeline}
\label{subsec:classical_ml}

Classical (shallow) machine learning requires a \emph{domain expert} to
manually engineer features before training a model.  The pipeline is:

\begin{equation}
  \label{eq:ml_pipeline}
  \underbrace{\text{Raw Input}}_{\mathbf{x}_\mathrm{raw}}
  \xrightarrow{\text{Feature Engineering}}
  \underbrace{\text{Feature Vector}}_{\boldsymbol{\phi}(\mathbf{x})}
  \xrightarrow{\text{Classifier / Regressor}}
  \hat{y}.
\end{equation}

The feature engineering step encodes domain knowledge: for image data this
might be HOG descriptors or SIFT keypoints; for audio it might be MFCC
coefficients; for text it might be TF-IDF bag-of-words vectors.

\begin{example}[title=Spam Email Detection (Rule-Based vs.\ ML)]
\textbf{Rule-based}: A programmer explicitly lists keywords (``free'',
``prize'', ``winner'') and directs the program to flag emails containing
them.  The programmer tells the system exactly what to look for.

\textbf{Classical ML}: The engineer labels a set of emails as
\texttt{spam=1} or \texttt{spam=0} and extracts features (word frequency
vectors, sender domain, etc.).  A logistic regression or SVM learns the
decision boundary from data, without being told which words matter.

\textbf{Deep Learning}: Raw email text (or pixels, if image-based) is fed
directly to a deep network.  The network simultaneously discovers relevant
features \emph{and} learns the classification rule end-to-end.
\end{example}

\subsection{Deep Learning Pipeline}
\label{subsec:dl_pipeline}

Deep Learning eliminates the manual feature-engineering step:

\begin{equation}
  \label{eq:dl_pipeline}
  \underbrace{\text{Raw Input}}_{\mathbf{x}_\mathrm{raw}}
  \xrightarrow{\text{Deep Network (end-to-end)}}
  \hat{y}.
\end{equation}

The early layers of the network automatically learn low-level features
(edges, textures, phonemes), and progressively deeper layers combine them
into higher-level semantic representations.  This is the key distinction:

\begin{definition}[Deep Learning]
\label{def:deep_learning}
\emph{Deep Learning} is a subfield of machine learning that uses multi-layer
neural networks (typically $L \geq 3$ hidden layers) to learn hierarchical
representations directly from raw data, obviating manual feature engineering.
The learned representation at layer $l$ is:
\begin{align}
  \label{eq:hierarchical}
  \mathbf{h}^{(1)} &= \phi\!\left(\mathbf{W}^{(1)}\mathbf{x} + \mathbf{b}^{(1)}\right), \\
  \mathbf{h}^{(l)} &= \phi\!\left(\mathbf{W}^{(l)}\mathbf{h}^{(l-1)} + \mathbf{b}^{(l)}\right),
                    \quad l = 2,\ldots,L, \\
  \hat{y}          &= g\!\left(\mathbf{W}^{(L+1)}\mathbf{h}^{(L)} + \mathbf{b}^{(L+1)}\right),
\end{align}
where $g(\cdot)$ is the output activation (e.g., softmax for classification).
\end{definition}

\subsection{Comparison Table: Shallow NN vs.\ Deep Learning}
\label{subsec:comparison}

The following table is reproduced from Slide V08 of the lecture:

\begin{table}[ht]
\centering
\caption{Neural Networks (shallow) vs.\ Deep Learning --- key distinctions (Slide V08).}
\label{tab:nn_vs_dl}
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Aspect} & \textbf{Neural Networks (Shallow)} & \textbf{Deep Learning} \\
\midrule
Depth & 1--2 hidden layers & Many hidden layers ($L \gg 2$) \\
\addlinespace
Training algorithm & Back-propagation works well & BP struggles; vanishing gradients \\
\addlinespace
Data requirement & Moderate dataset sufficient & Large datasets required \\
\addlinespace
Feature engineering & Manual features necessary & Automatic feature learning \\
\addlinespace
Model size & Small parameter count & Millions to billions of parameters \\
\addlinespace
Computational cost & Low (CPU-feasible) & High (GPU/TPU required) \\
\addlinespace
Representative models & Perceptron, MLP, SVM & CNN, LSTM, Transformer, GPT \\
\bottomrule
\end{tabularx}
\end{table}

\begin{keyinsight}
The fundamental reason shallow networks needed manual features is their
\emph{limited capacity}: with only 1--2 hidden layers, the model lacks the
representational power to discover complex hierarchical features itself.
Depth provides exponential representational advantage, a fact formalised
by the depth separation theorems in circuit complexity theory.
\end{keyinsight}

%% ============================================================
\section{The AI Taxonomy: AI $\supset$ ML $\supset$ NN $\supset$ DL $\supset$ GAI}
\label{sec:taxonomy}
%% ============================================================

\subsection{The Nested Hierarchy}
\label{subsec:nested}

The final slides of the session situate Deep Learning within the broader
intellectual landscape using a nested-circles diagram (Slide V10).

\begin{figure}[ht]
\centering
\begin{tikzpicture}[scale=1.0]
  %% Outermost: AI
  \draw[thick, fill=cyan!8]
    (0,0) ellipse (6.2cm and 4.2cm);
  \node[font=\large\bfseries] at (0,3.5) {Artificial Intelligence (AI)};

  %% ML
  \draw[thick, fill=blue!10]
    (0,-0.2) ellipse (4.8cm and 3.1cm);
  \node[font=\bfseries] at (0,2.3) {Machine Learning (ML)};

  %% NN
  \draw[thick, fill=green!12]
    (0,-0.5) ellipse (3.4cm and 2.0cm);
  \node[font=\bfseries] at (0,1.1) {Neural Networks (NN)};

  %% DL
  \draw[thick, fill=orange!18]
    (0,-0.7) ellipse (2.0cm and 1.1cm);
  \node[font=\small\bfseries] at (0,0.1) {Deep Learning (DL)};

  %% GAI
  \draw[thick, fill=red!20]
    (0,-0.9) ellipse (0.9cm and 0.5cm);
  \node[font=\tiny\bfseries] at (0,-0.9) {GAI};
\end{tikzpicture}
\caption{The AI taxonomy: nested circles showing AI $\supset$ ML $\supset$ NN
$\supset$ DL $\supset$ GAI (Slide V10).
GAI = Generative AI.  The course covers the three innermost regions.}
\label{fig:taxonomy}
\end{figure}

\subsection{Definitions of Each Level}
\label{subsec:level_definitions}

\begin{definition}[Artificial Intelligence]
\label{def:ai}
\emph{Artificial Intelligence} (AI) encompasses all methods developed to
enable computing machines to perform tasks that require human-like
intelligence: discovery, learning, reasoning under uncertainty, creativity,
and handling novel situations.  It includes both \emph{symbolic} (rule-based,
logic-driven) and \emph{connectionist} (neural-network-based) approaches.
\end{definition}

\begin{definition}[Machine Learning]
\label{def:ml}
\emph{Machine Learning} (ML) is the subfield of AI concerned with systems
that learn from data.  Rather than being explicitly programmed with rules, an
ML model discovers the mapping $f: \mathcal{X} \to \mathcal{Y}$ from a
training dataset $\{(\mathbf{x}_i, y_i)\}_{i=1}^N$ by minimising a loss
functional $\mathcal{L}$ (see \cref{eq:loss}).
\end{definition}

\begin{definition}[Neural Networks]
\label{def:nn}
\emph{Neural Networks} (NN) are a subset of ML models inspired by the
biological brain.  They are parameterised computation graphs of artificial
neurons arranged in layers (see \cref{def:ann}).  The classical shallow
neural network with 1--2 hidden layers is the precursor to deep learning.
\end{definition}

\begin{definition}[Generative AI]
\label{def:gai}
\emph{Generative AI} (GAI) denotes deep learning models trained to generate
new data samples (images, text, audio) from a learned distribution.
Representative architectures include Variational Autoencoders (VAEs),
Generative Adversarial Networks (GANs), and large language models based on
the Transformer (GPT, Gemini, Claude, etc.).
\end{definition}

\begin{remark}
\label{rem:connectionism}
The AI discussed throughout this course is \emph{connectionist AI} — neural
network and deep-learning-based.  The alternative tradition of
\emph{classical/symbolic AI} (expert systems, Prolog-style reasoning,
knowledge representation via symbols and rules) is not covered in this course.
\end{remark}

%% ============================================================
\section{Architecture Diagrams}
\label{sec:diagrams}
%% ============================================================

\subsection{Classical ML vs.\ Deep Learning Pipeline}
\label{subsec:ml_dl_diagram}

\cref{fig:ml_vs_dl} recreates Slide V09 showing the fundamental difference
between the two pipelines.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    box/.style={draw, rounded corners, fill=#1, minimum width=2.6cm,
                minimum height=0.9cm, align=center, font=\small},
    arr/.style={-{Stealth[length=6pt]}, thick},
    every node/.style={font=\small}
  ]

  %% ---- Classical Machine Learning row ----
  \node[box=gray!20]  (input_ml)  at (0,0)      {Raw\\Input};
  \node[box=yellow!35](feat_ml)   at (3.4,0)    {Feature\\Engineering\\(manual)};
  \node[box=blue!20]  (model_ml)  at (7.2,0)    {Classical\\Classifier\\/ Regressor};
  \node[box=green!25] (out_ml)    at (10.6,0)   {Output\\$\hat{y}$};

  \draw[arr] (input_ml)  -- (feat_ml);
  \draw[arr] (feat_ml)   -- (model_ml);
  \draw[arr] (model_ml)  -- (out_ml);

  \node[font=\small\bfseries, anchor=east] at (-0.6,0) {ML:};

  %% ---- Deep Learning row ----
  \node[box=gray!20]  (input_dl)  at (0,-2.4)  {Raw\\Input};
  \node[box=orange!25](dl_block)  at (4.2,-2.4){Deep Neural Network\\(feature extraction\\+ classification)};
  \node[box=green!25] (out_dl)    at (8.6,-2.4){Output\\$\hat{y}$};

  \draw[arr] (input_dl)  -- (dl_block);
  \draw[arr] (dl_block)  -- (out_dl);

  \node[font=\small\bfseries, anchor=east] at (-0.6,-2.4) {DL:};

  %% ---- annotation braces ----
  \draw[decorate,decoration={brace,mirror,amplitude=5pt},thick]
    (3.0,-0.6) -- (7.6,-0.6)
    node[midway, below=6pt, font=\footnotesize\itshape]
      {domain expert required};

  \draw[decorate,decoration={brace,mirror,amplitude=5pt},thick]
    (1.5,-3.0) -- (7.2,-3.0)
    node[midway, below=6pt, font=\footnotesize\itshape]
      {learned end-to-end};

\end{tikzpicture}
\caption{Classical ML vs.\ Deep Learning pipelines (Slide V09).
In classical ML the feature engineering block is separate and requires
domain expertise; in deep learning it is subsumed within the network and
learned automatically from data.}
\label{fig:ml_vs_dl}
\end{figure}

\subsection{A Simple Feedforward Neural Network}
\label{subsec:ffnn_diagram}

\cref{fig:ffnn} illustrates a two-hidden-layer feedforward network, the type
built in the Week~1 hands-on session.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    neuron/.style={circle, draw, fill=#1, minimum size=0.75cm, inner sep=0pt},
    arr/.style={-{Stealth[length=5pt]}, gray!70},
    lbl/.style={font=\scriptsize\bfseries, align=center}
  ]

  \def\layersep{2.6cm}

  %% Input layer (4 nodes)
  \foreach \y/\name in {1.5/x_1, 0.5/x_2, -0.5/x_3, -1.5/x_4}{
    \node[neuron=cyan!25] (I\name) at (0,\y) {};
    \node[font=\tiny] at (0,\y) {$\name$};
  }

  %% Hidden layer 1 (5 nodes)
  \foreach \y in {2, 1, 0, -1, -2}{
    \node[neuron=yellow!35] (H1-\y) at (\layersep,\y) {};
  }

  %% Hidden layer 2 (4 nodes)
  \foreach \y in {1.5, 0.5, -0.5, -1.5}{
    \node[neuron=orange!35] (H2-\y) at (2*\layersep,\y) {};
  }

  %% Output layer (3 nodes)
  \foreach \y/\name in {1/y_1, 0/y_2, -1/y_3}{
    \node[neuron=green!35] (O\name) at (3*\layersep,\y) {};
    \node[font=\tiny] at (3*\layersep,\y) {$\name$};
  }

  %% Connections: input -> H1
  \foreach \src in {x_1,x_2,x_3,x_4}{
    \foreach \dst in {2,1,0,-1,-2}{
      \draw[arr] (I\src) -- (H1-\dst);
    }
  }

  %% Connections: H1 -> H2
  \foreach \src in {2,1,0,-1,-2}{
    \foreach \dst in {1.5,0.5,-0.5,-1.5}{
      \draw[arr] (H1-\src) -- (H2-\dst);
    }
  }

  %% Connections: H2 -> output
  \foreach \src in {1.5,0.5,-0.5,-1.5}{
    \foreach \dst in {y_1,y_2,y_3}{
      \draw[arr] (H2-\src) -- (O\dst);
    }
  }

  %% Layer labels
  \node[lbl] at (0,-2.8)         {Input\\Layer};
  \node[lbl] at (\layersep,-2.8) {Hidden\\Layer 1};
  \node[lbl] at (2*\layersep,-2.8){Hidden\\Layer 2};
  \node[lbl] at (3*\layersep,-2.8){Output\\Layer};

\end{tikzpicture}
\caption{A fully-connected feedforward neural network with two hidden layers
(4-5-4-3 architecture).  This is the architecture built in the Week~1
hands-on session.  Each arrow represents a learnable weight $w_{jk}$.}
\label{fig:ffnn}
\end{figure}

%% ============================================================
\section{Key Concepts: Backpropagation and Gradient Descent}
\label{sec:backprop}
%% ============================================================

\subsection{Gradient Descent}
\label{subsec:gd}

The standard parameter update rule is stochastic gradient descent (SGD):
\begin{equation}
  \label{eq:sgd}
  \boldsymbol{\theta}^{(t+1)}
  = \boldsymbol{\theta}^{(t)}
  - \eta \, \nabla_{\boldsymbol{\theta}} \mathcal{L}\!\left(\boldsymbol{\theta}^{(t)}\right),
\end{equation}
where $\eta > 0$ is the learning rate and $\nabla_{\boldsymbol{\theta}}
\mathcal{L}$ is the gradient of the loss with respect to the parameters.

\subsection{The Backpropagation Algorithm}
\label{subsec:bp_algo}

Backpropagation efficiently computes $\nabla_{\boldsymbol{\theta}} \mathcal{L}$
by a single forward pass (to cache activations) followed by a single backward
pass (to propagate error signals).

\begin{algorithm}[H]
\caption{Backpropagation for a fully-connected network}
\label{alg:backprop}
\SetAlgoLined
\KwIn{Training pair $(\mathbf{x}, y)$; weights $\{(\mathbf{W}^{(l)}, \mathbf{b}^{(l)})\}_{l=1}^{L+1}$;
      learning rate $\eta$; activation $\phi$}
\KwOut{Updated weights}

\BlankLine
\tcp{--- Forward Pass ---}
$\mathbf{a}^{(0)} \leftarrow \mathbf{x}$\;
\For{$l = 1$ \KwTo $L+1$}{
  $\mathbf{z}^{(l)} \leftarrow \mathbf{W}^{(l)}\mathbf{a}^{(l-1)} + \mathbf{b}^{(l)}$\;
  $\mathbf{a}^{(l)} \leftarrow \phi(\mathbf{z}^{(l)})$ \tcp*{element-wise}
}
Compute loss $\mathcal{L} = \ell(y, \mathbf{a}^{(L+1)})$\;

\BlankLine
\tcp{--- Backward Pass ---}
$\boldsymbol{\delta}^{(L+1)} \leftarrow \nabla_{\mathbf{a}^{(L+1)}}\mathcal{L}
                                \odot \phi'(\mathbf{z}^{(L+1)})$\;
\For{$l = L$ \KwTo $1$}{
  $\boldsymbol{\delta}^{(l)} \leftarrow
    \left(\mathbf{W}^{(l+1)\top}\boldsymbol{\delta}^{(l+1)}\right)
    \odot \phi'(\mathbf{z}^{(l)})$\;
}

\BlankLine
\tcp{--- Weight Update ---}
\For{$l = 1$ \KwTo $L+1$}{
  $\mathbf{W}^{(l)} \leftarrow \mathbf{W}^{(l)}
    - \eta\, \boldsymbol{\delta}^{(l)} {\mathbf{a}^{(l-1)}}^{\!\top}$\;
  $\mathbf{b}^{(l)} \leftarrow \mathbf{b}^{(l)}
    - \eta\, \boldsymbol{\delta}^{(l)}$\;
}
\Return Updated weights\;
\end{algorithm}

\begin{keyinsight}[title=Why BP Fails in Deep Networks]
In \cref{alg:backprop}, the backward-pass loop multiplies error signals
by $\mathbf{W}^{(l)\top}$ at each layer.  If the spectral norm
$\|\mathbf{W}^{(l)}\|_2 < 1$ (common with random initialisation) or
$|\phi'(z)| \ll 1$ (sigmoid saturation), the signals in early layers
vanish exponentially with depth.  This is the \emph{vanishing gradient
problem} that blocked progress until the 2006 Hinton breakthrough and the
subsequent adoption of ReLU activations ($\phi(z) = \max(0,z)$, derivative
$= 1$ for $z>0$).
\end{keyinsight}

\subsection{Greedy Layer-Wise Pre-Training}
\label{subsec:greedy}

The algorithmic insight of Hinton \emph{et al.}\ (2006) is to decompose the
deep training problem into a sequence of shallow problems:

\begin{enumerate}
  \item Train Layer~1 as an RBM on the raw data $\{\mathbf{x}_i\}$.
  \item Use the hidden representations from Layer~1 as inputs to train
        Layer~2 as a second RBM.
  \item Continue greedily for all $L$ layers.
  \item Use the stacked pre-trained weights to initialise the full deep
        network, then fine-tune end-to-end with BP.
\end{enumerate}

The energy function of an RBM is:
\begin{equation}
  \label{eq:rbm_energy}
  E(\mathbf{v}, \mathbf{h}) = -\mathbf{b}^\top \mathbf{v}
                              - \mathbf{c}^\top \mathbf{h}
                              - \mathbf{v}^\top \mathbf{W} \mathbf{h},
\end{equation}
where $\mathbf{v}$ are visible units, $\mathbf{h}$ are hidden units,
$\mathbf{b}$, $\mathbf{c}$ are biases, and $\mathbf{W}$ are the RBM weights.
The joint distribution is:
\begin{equation}
  \label{eq:rbm_joint}
  P(\mathbf{v},\mathbf{h}) = \frac{1}{Z}\exp\!\left(-E(\mathbf{v},\mathbf{h})\right),
  \qquad Z = \sum_{\mathbf{v},\mathbf{h}} \exp\!\left(-E(\mathbf{v},\mathbf{h})\right).
\end{equation}

%% ============================================================
\section{Why Sequence Models? Motivating Example}
\label{sec:seqmodels}
%% ============================================================

Standard feedforward networks and CNNs process inputs as fixed-size
\emph{static} feature vectors; they cannot model the order in which elements
appear.  The instructor motivates sequence models with a concrete example:

\begin{example}[title=Mobile Phone Number Recognition]
All Indian mobile numbers are 10-digit strings drawn from the alphabet
$\{0,1,\ldots,9\}$.  Two different numbers, say $9876543210$ and $9801234567$,
use exactly the same digit set.  Their \emph{distribution} of digits
is similar, but their \emph{sequential ordering} is completely different.

A feedforward classifier operating on a bag-of-digits feature (histogram of
digit occurrences) would be unable to distinguish the two numbers.  Only a
model that respects the temporal order — a Recurrent Neural Network —
can capture this discriminatory information.
\end{example}

\begin{keyinsight}
This motivating example generalises widely: speech (phoneme sequences),
text (word sequences), DNA (nucleotide sequences), financial time-series
(price sequences) — in all of these, the \emph{order matters}.
Recurrent Neural Networks (RNNs), LSTMs, GRUs, and Transformers were all
developed to exploit this sequential structure.
\end{keyinsight}

Formally, a vanilla RNN processes a sequence $x_1, x_2, \ldots, x_T$ by
maintaining a hidden state $\mathbf{h}_t$:
\begin{align}
  \label{eq:rnn}
  \mathbf{h}_t &= \phi\!\left(\mathbf{W}_h\, \mathbf{h}_{t-1}
                              + \mathbf{W}_x\, \mathbf{x}_t
                              + \mathbf{b}\right), \\
  \mathbf{o}_t &= \mathbf{V}\, \mathbf{h}_t,
\end{align}
where $\mathbf{W}_h$, $\mathbf{W}_x$, $\mathbf{V}$, $\mathbf{b}$ are shared
parameters across all time steps $t$.

%% ============================================================
\section{Summary and Looking Ahead}
\label{sec:summary}
%% ============================================================

\subsection{Session Summary}
\label{subsec:session_summary}

Session~01 served as a complete orientation to the Deep Learning course.
The key take-home messages are:

\begin{enumerate}[label=\textbf{T\arabic*.}]
  \item \textbf{Course design}: The course is hands-on and PyTorch-driven,
        spanning perceptrons (Week~1) through Transformers and Generative
        Models (Weeks~8--13).
  \item \textbf{The 2006 breakthrough}: Geoffrey Hinton \emph{et al.}\ showed
        for the first time that a deep neural network could be reliably trained
        and that doing so yields state-of-the-art performance.  This paper
        re-ignited the field and is the root of modern AI.
  \item \textbf{Classical ML vs.\ Deep Learning}: Classical ML uses manual
        feature engineering followed by a shallow model; Deep Learning learns
        features and classification jointly, end-to-end, from raw data.
  \item \textbf{The AI hierarchy}: AI $\supset$ ML $\supset$ NN $\supset$ DL
        $\supset$ GAI.  This course focuses on NN, DL, and GAI.
  \item \textbf{Three pillars of success}: algorithmic advances
        (pre-training, ReLU, normalisation), big data, and GPU computing.
  \item \textbf{Sequence models}: RNNs and their variants (LSTM, GRU,
        Transformer) are necessary when discriminatory information lies in
        the \emph{order} of inputs, not just their distribution.
\end{enumerate}

\subsection{What the Student Should Know After This Course}
\label{subsec:takeaway}

Quoting the instructor verbatim:
\begin{quote}
``If you have done the deep learning course, you should be very comfortable
with Convolutional Networks.  You should be very comfortable with LSTM.
You should have very good working experience with Transformers and Generative
Adversarial Networks.  That should be the take-home message for all of you
when you have gone through this course.''
--- Prof.\ S R M Prasanna
\end{quote}

\subsection{Connections to Upcoming Sessions}
\label{subsec:upcoming}

\begin{table}[ht]
\centering
\caption{Road map from Session~01 to the rest of the course.}
\label{tab:roadmap}
\begin{tabular}{lll}
\toprule
\textbf{Phase} & \textbf{Sessions} & \textbf{Core Concept Introduced} \\
\midrule
Foundations    & 1--2  & Perceptron, BP, 2-layer NN \\
Deep NNs       & 3--4  & Deep MLP, vanishing gradients, pre-training \\
Spatial models & 5--8  & Convolutional networks, pooling, Grad-CAM \\
Sequence models& 9--12 & RNN, BPTT, LSTM, GRU \\
Attention      & 13--14 & Seq2Seq, Bahdanau attention, Transformer \\
Generative     & 15--18 & VAE, GAN, GPT \\
\bottomrule
\end{tabular}
\end{table}

\begin{keyinsight}[title=The Big Picture]
The entire curriculum can be understood through one lens: each successive
architecture relaxes a bottleneck of its predecessor.
\begin{itemize}
  \item Perceptron $\to$ MLP: adds non-linearity and hidden representations.
  \item MLP $\to$ Deep MLP: adds depth for hierarchical abstraction.
  \item Deep MLP $\to$ CNN: adds spatial inductive bias (weight sharing,
        local connectivity).
  \item CNN/MLP $\to$ RNN: adds temporal inductive bias (recurrence).
  \item RNN $\to$ LSTM: adds gating to solve the long-range dependency problem.
  \item RNN $\to$ Transformer: replaces recurrence with self-attention for
        parallelisable, global-context modelling.
  \item Transformer $\to$ GAI models: repurposes the Transformer as a
        conditional generative model.
\end{itemize}
\end{keyinsight}

\newpage
%% ============================================================
%%  Appendix
%% ============================================================
\appendix

\section{Activation Functions Reference}
\label{app:activations}

\begin{table}[ht]
\centering
\caption{Common activation functions used in deep networks.}
\label{tab:activations}
\begin{tabular}{llll}
\toprule
\textbf{Name} & \textbf{Formula} $\phi(z)$ & \textbf{Derivative} $\phi'(z)$ & \textbf{Range} \\
\midrule
Sigmoid   & $\dfrac{1}{1+e^{-z}}$          & $\phi(z)(1-\phi(z))$          & $(0,1)$ \\[8pt]
Tanh      & $\dfrac{e^z - e^{-z}}{e^z+e^{-z}}$ & $1-\phi(z)^2$            & $(-1,1)$ \\[8pt]
ReLU      & $\max(0,z)$                     & $\mathbf{1}[z>0]$             & $[0,\infty)$ \\[8pt]
Leaky ReLU & $\max(\alpha z,z)$, $\alpha\ll 1$ & $1$ if $z>0$, else $\alpha$ & $(-\infty,\infty)$ \\[8pt]
Softmax   & $\dfrac{e^{z_k}}{\sum_j e^{z_j}}$ & (Jacobian matrix)          & $(0,1)$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{remark}
\label{rem:relu}
The switch from sigmoid/tanh to ReLU activations was one of the most
impactful practical innovations of the 2010s.  Because $\phi'(z)=1$ for
$z>0$, ReLU does not suppress gradients in the backward pass, substantially
mitigating the vanishing gradient problem (see \cref{eq:vanishing}).
\end{remark}

\section{Glossary}
\label{app:glossary}

\begin{description}[style=nextline]
  \item[Artificial Intelligence (AI)]
    The broad field of creating computing systems that mimic human
    intelligence.

  \item[Machine Learning (ML)]
    A subfield of AI where systems learn from data without being explicitly
    programmed.

  \item[Neural Network (NN)]
    A layered computational graph of artificial neurons; the precursor to
    deep learning.

  \item[Deep Learning (DL)]
    ML with multi-layer neural networks that learn hierarchical
    representations from raw data.

  \item[Generative AI (GAI)]
    Deep learning models that generate new data samples (text, images,
    audio) from a learned distribution.

  \item[Backpropagation (BP)]
    Algorithm for computing gradients of the loss through a neural network
    using the chain rule of calculus.

  \item[Vanishing Gradient]
    The exponential decay of gradient signals in early layers of a deep
    network, caused by repeated multiplication by small derivative values.

  \item[Greedy Layer-wise Pre-training]
    A technique introduced by Hinton \emph{et al.}\ (2006) for training deep
    networks by successively training each layer as an RBM before global
    fine-tuning.

  \item[Restricted Boltzmann Machine (RBM)]
    A two-layer energy-based generative model used as the building block
    of Deep Belief Networks.

  \item[Convolutional Neural Network (CNN)]
    A deep network with weight-shared, locally-connected filters designed
    for spatially-structured inputs (images).

  \item[Recurrent Neural Network (RNN)]
    A network with cyclic connections that maintains a hidden state over
    time steps, enabling sequence modelling.

  \item[Long Short-Term Memory (LSTM)]
    An RNN variant with multiplicative gating mechanisms that mitigate
    the vanishing gradient problem in long sequences.

  \item[Transformer]
    An architecture based entirely on self-attention (no recurrence) that
    models global dependencies in sequences.  Foundation of GPT and BERT.

  \item[Variational Autoencoder (VAE)]
    A generative model that learns a latent-space distribution and can
    sample new data by decoding latent vectors.

  \item[Generative Adversarial Network (GAN)]
    A generative model consisting of a Generator and a Discriminator
    trained adversarially.

  \item[PyTorch]
    An open-source deep learning framework developed by Meta AI.  Used
    throughout this course for hands-on implementations.
\end{description}

\end{document}
